---
title: "Programming Assignment 1"
author: "suveyda"
date: "January 18, 2017"
output:
  html_document: default
  pdf_document: default
---

# Programming Assignment 1: Air Pollution

## R Programming by Johns Hopkins University

### Introduction

For this programming assignment we need to write three functions that are meant to interact with dataset that accompanies this assignment. 

### Data

The dataset is contained in a zip file specdata.zip that you can download from the Coursera web site. The zip file contains 332 comma-separated-value (CSV) files containing pollution monitoring data for fine particulate matter (PM) air pollution at 332 locations in the United States. Each file contains data from a single monitor and the ID number for each monitor is contained in the file name. For example, data for monitor 200 is contained in the file "200.csv". Each file contains three variables:

* Date: the date of the observation in YYYY-MM-DD format (year-month-day)
* sulfate: the level of sulfate PM in the air on that date (measured in micrograms per cubic meter)
* nitrate: the level of nitrate PM in the air on that date (measured in micrograms per cubic meter)

In each file there are many days where either sulfate or nitrate (or both) are missing (coded as NA). This is common with air pollution monitoring data in the United States.

As mentioned above, the folder contains 332 files named as:

* 001.csv, 002.csv, ..., 009.csv
* 010.csv, 011.csv, ..., 099.csv
* 100.csv, 101.csv, ..., 332.csv

In order to process these files, the below 'filePath' function is written.

```{r filePath}
filePath <- function(directory, x)
{
	if(x < 10)
		paste(directory, "/00", x, ".csv", sep = "")
	else if(x < 100)
		paste(directory, "/0", x, ".csv", sep = "")
	else
		paste(directory, "/", x, ".csv", sep = "")
}
```

### Part 1: Pollutantmean

In the first part, we write a function named **'pollutantmean'** that calculates the mean of a pollutant (sulfate or nitrate) across a specified list of monitors. The function 'pollutantmean' takes three arguments: 'directory', 'pollutant', and 'id'. Given a vector monitor ID numbers, 'pollutantmean' reads that monitors' particulate matter data from the directory specified in the 'directory' argument and returns the mean of the pollutant across all of the monitors, ignoring any missing values coded as NA. 

```{r pollutantmean}
pollutantmean <- function(directory, pollutant, id = 1:332)
{
	all <- vector(mode="numeric", length=0)
	
	for(i in seq_along(id)){
		file_path = filePath(directory, id[i])
		
		data <- read.csv(file_path, header = TRUE, sep = ",")
		
		if(pollutant == "sulfate")
			sub_data <- data$sulfate[!is.na(data$sulfate)]
		else if(pollutant == "nitrate")
			sub_data <- data$nitrate[!is.na(data$nitrate)]
		else
			break
			
		all <- c(all, sub_data)
	}
	mean(all)
}
```

```{r}
pollutantmean("specdata", "sulfate", 1:10)
pollutantmean("specdata", "nitrate", 70:72)
pollutantmean("specdata", "nitrate", 23)
pollutantmean("specdata", "sulfate", 34)
pollutantmean("specdata", "nitrate")
```

### Part 2: Complete

In the second part of the assignment, we write a function that reads a directory full of files and reports the number of completely observed cases in each data file. The function returns a data frame where the first column is the name of the file and the second column is the number of complete cases. 

```{r complete}
complete <- function(directory, id = 1:332)
{
	df <- data.frame("id"=numeric(), "nobs"=numeric())
	
	for(i in seq_along(id)){
		file_path = filePath(directory, id[i])
		
		data <- read.csv(file_path, header = TRUE, sep = ",")
		count <- sum(complete.cases(data$sulfate, data$nitrate))
		
		df <- rbind(df, c(id[i], count))
	}
	colnames(df) <- c("id","nobs")
	df
}
```

```{r}
complete("specdata", 1)
complete("specdata", c(2, 4, 8, 10, 12))
complete("specdata", 30:25)
complete("specdata", 3)
complete("specdata", c(6, 10, 20, 34, 100, 200, 310))
complete("specdata", 54)
```

```{r}
set.seed(42)
cc <- complete("specdata", 332:1)
use <- sample(332, 10)
print(cc[use, "nobs"])
```


### Part 3: Corr

In this last part of the assignment, we write a function that takes a directory of data files and a threshold for complete cases and calculates the correlation between sulfate and nitrate for monitor locations where the number of completely observed cases (on all variables) is greater than the threshold. The function returns a vector of correlations for the monitors that meet the threshold requirement. If no monitors meet the threshold requirement, then the function returns a numeric vector of length 0. 

```{r corr}
corr <- function(directory, threshold = 0)
{
	id = 1:332
	correlations <- vector(mode="numeric", length=0)
	
	for(i in seq_along(id)){
		file_path = filePath(directory, id[i])

		data <- read.csv(file_path, header = TRUE, sep = ",")
		complete_data <- complete.cases(data$sulfate, data$nitrate)
		
		if(sum(complete_data) > threshold)
		{
			correlation <- cor(data$sulfate[complete_data], data$nitrate[complete_data])
			correlations <- c(correlations, correlation)
		}
	}
	correlations
}
```

```{r}
cr <- corr("specdata", 150)
head(cr)
summary(cr)

cr <- corr("specdata", 400)
head(cr)
summary(cr)

cr <- corr("specdata", 5000)
summary(cr)
length(cr)

cr <- corr("specdata")
summary(cr)
length(cr)

cr <- corr("specdata")                
cr <- sort(cr)                
set.seed(868)                
out <- round(cr[sample(length(cr), 5)], 4)
print(out)

cr <- corr("specdata", 129)                
cr <- sort(cr)                
n <- length(cr)                
set.seed(197)                
out <- c(n, round(cr[sample(n, 5)], 4))
print(out)

cr <- corr("specdata", 2000)                
n <- length(cr)                
cr <- corr("specdata", 1000)                
cr <- sort(cr)
print(c(n, round(cr, 4)))
```

