---
title: "Programming_Assignment_2"
author: "suveyda"
date: "January 18, 2017"
output: html_document
---
# Programming Assignment 2: Lexical Scoping

## R Programming by Johns Hopkins University

### Introduction

In this programming assignment we need to write an R function that is able to cache potentially time-consuming computations. For example, taking the mean of a numeric vector is typically a fast operation. However, for a very long vector, it may take too long to compute the mean, especially if it has to be computed repeatedly (e.g. in a loop). If the contents of a vector are not changing, it may make sense to cache the value of the mean so that when we need it again, it can be looked up in the cache rather than recomputed.

### Caching the Inverse of a Matrix

Matrix inversion is usually a costly computation and there may be some benefit to caching the inverse of a matrix rather than compute it repeatedly. The requirement of this assignment is to write a pair of functions that cache the inverse of a matrix.

### Function 1: makeCacheMatrix

This function creates a special "matrix" object that can cache its inverse.

```{r makeCacheMatrix}
makeCacheMatrix <- function(x = matrix()) {
        inverse_x <- NULL
        set <- function(y) {
                x <<- y
                inverse_x <<- NULL
        }
        get <- function() x
        set_inverse <- function(matrix) inverse_x <<- matrix
        get_inverse <- function() inverse_x
        list(set = set, get = get,
             set_inverse = set_inverse,
             get_inverse = get_inverse)
}
```

### Function 2: cacheSolve

This function computes the inverse of the special "matrix" returned by makeCacheMatrix above. If the inverse has already been calculated (and the matrix has not changed), then the cachesolve will retrieve the inverse from the cache.

```{r cacheSolve}
cacheSolve <- function(x, ...) {
        inverse_x <- x$get_inverse()
        if(!is.null(inverse_x)) {
                message("getting cached data")
                return(inverse_x)
        }
        data <- x$get()
        inverse_x <- solve(data, ...)
        x$set_inverse(inverse_x)
        inverse_x
}
```

Here is a function that can test the above functions:

```{r}
test = function(matrix){
    temp = makeCacheMatrix(matrix)
        
    start.time = Sys.time()
    cacheSolve(temp)
    dur = Sys.time() - start.time
    print(dur)
        
    start.time = Sys.time()
    cacheSolve(temp)
    dur = Sys.time() - start.time
    print(dur)
}
```

```{r}
set.seed(1110201)
r = rnorm(1000000)
matrix = matrix(r, nrow=1000, ncol=1000)
test(matrix)
```

As you can see from the above example, the processing time decreased vastly due to caching.